---
title: "dplyr tutorial"
author: "James Ball"
date: "27/07/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# dplyr intro

## dplyr basics

First we'll look at three key `dplyr` functions for basic data manipulation:

- Pick observations by their values with `filter()`
- Create new variables with functions of existing variables with `mutate()`.
- Collapse many values down to a single summary `summarize()`.

These can all be used in conjunction with `group_by()` which changes the scope of each function from operating on the entire dataset to operating on it group-by-group.

All `dplyr` verbs work similarly:

1. The first argument is a data frame
2. The subsequent arguments describe what to do with the data frame
3. The result is a new data frame

See the [data transformation cheatsheet](https://github.com/rstudio/cheatsheets/blob/master/data-transformation.pdf) for help.

we will use the `dplyr` package and example data from `nycflights13`

```{r libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
#install.packages("nycflights13")
library(nycflights13)
```

Take a look at the data

```{r data}
head(flights)
```

Q1/ Can you use `filter()` to inspect flights from the first day of the year? (hint: use `month == ` and `day == `)?
(Use a pipe to pass the data into the filter function)
```{r month, echo=FALSE}
flights %>% 
  filter(month == 1, day == 1)
```


To filter the flights for November and December, you can use the `|` ("or") logical operator

```{r or}
flights %>% 
  filter(month == 11 | month == 12)
```

Alternatively, it can be useful to use the `%in%` operator

```{r in}
flights %>% 
  filter(month %in% c(11, 12))
```

## New variables with `mutate()`

To create a more managable Tibble we will use the `select()` function to choose the variables that are of interest to us. Note the use of `ends_withs()` to select multiple relevant variables.

```{r select}
flights_sml <- 
  flights %>% 
  select( 
    year:day, 
    ends_with("delay"), 
    distance, 
    air_time
  )
flights_sml
```

Now we would like to create some new variables

```{r gain}
flights_sml %>% 
  mutate(
    gain = arr_delay - dep_delay
  )
```

Here we have created the variable `gain` which tells us the difference between the arrival delay and the departure delay.

Q2/ How would you create a new variable that tells us the average speed during the flight (`speed`)

```{r speed, echo=FALSE}
flights_sml %>% 
  mutate(
    gain = arr_delay - dep_delay,
    speed = distance / air_time * 60
  )
```

## Grouped summaries

The verb `summarize()` collapses a data frame to a single row. For examples, we can use it to calculate the average delay.

```{r delay}
flights %>% 
  summarize(delay = mean(dep_delay, na.rm = TRUE))
```

We can combine it with `group_by()` which changes the unit of analysis from the complete dataset to individual groups.

How would you create a data frame that tells you the average delay in each month? (hint: use two pipe operations)

```{r delaymonth, echo=FALSE}
flights %>% 
  group_by(year, month) %>% 
  summarize(delay = mean(dep_delay, na.rm = TRUE))
```

## Extension exercise

Imagine that we want to explore the relationship between the distance and average delay for each location.

To prepare this data we can:

1. Use `group_by()` to group flights by destination.

2. Use `summarize()` to count the flights for each destination and calculate the mean distance and arrival delay e.g.
```
summarize(
  by_dest,
  count = n(),
  dist = mean(distance, na.rm = TRUE),
  delay = mean(arr_delay, na.rm = TRUE)
)
```

3. Filter to remove noisy points and Honolulu airport, which is almost twice as far away as the next closest airport.
```
delay <- filter(delay, count > 20, dest != "HNL")
```

We can then use ggplot to display these results.

How can we combine these operations using pipes (`%>%`)?


```{r plot, echo=FALSE, message=FALSE}
delays <- 
  flights %>% 
  group_by(dest) %>% 
  summarize(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>% 
  filter(count > 20, dest != "HNL")
ggplot(data = delays, mapping = aes(x = dist, y = delay)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)
```


```{r plot2, echo=FALSE, eval=FALSE}
flights %>% 
  group_by(dest) %>% 
  summarize(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>% 
  filter(count > 20, dest != "HNL") %>%
  ggplot(mapping = aes(x = dist, y = delay)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)
```

See [Data Challenge Lab](https://dcl-wrangle.stanford.edu/) for solutions.