---
title: "tidyverse tutorial"
author: "James Ball"
date: "26/07/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Tidy data and data wrangling

We will need the `tidyr`, `dplyr` and `readr` packages. We will use example data
from the `dcldata` package.

```{r packages, warning=FALSE, message=FALSE}
library(tidyr)
library(dplyr)
library(readr)
#devtools::install_github("dcl-docs/dcldata")
library(dcldata) # for the example datasets
```

## Pivoting

We will be using `tidyr`'s the `pivot_longer` and `pivot_wider` functions rather
than the outdated `spread` and `gather`.

We will be using the `example_eagle_nests` dataset which is loaded as a Tibble.

```{r data}
example_eagle_nests
```

Q1/
Is this data tidy? Give reasons

Q2/
Go through these steps, and use the `pivot_longer` function and a pipe to define
a tidy version of the dataset:

1. What are the variables?
2. Which columns should we pivot?
3. Which columns contain the values?
4. Name the column that will store the column values.


```{r pivot_long, eval=FALSE}
example_eagle_nests_tidy <-   
  example_eagle_nests %>% 
    pivot_longer(
      # Step 2
      # Step 3
      # Step 4
    )
```
```{r pivot_long_run, echo=FALSE}
example_eagle_nests_tidy <-
  example_eagle_nests %>% 
    pivot_longer(
      cols = c(`2007`, `2009`), 
      names_to = "year", 
      values_to = "num_nests"
  )
example_eagle_nests_tidy
```

`year` is returned as character column.

Q3/ Can you use the `names_ptypes` arguement to return it as a more useful vector type?

```{r names, echo=FALSE}
example_eagle_nests %>% 
  pivot_longer(
    cols = c(`2007`, `2009`), 
    names_to = "year", 
    names_ptypes = list(year = integer()),
    values_to = "num_nests"
  )
```

`pivot_longer()` also has a values_ptypes argument that controls the type of the values_to column. `values_ptypes` is specified in the same way as `names_ptypes`.

Q4/ Using `pivot_longer()`, how would we change `num_nests` from its default type (double) to an integer?

```{r values, echo=FALSE}
example_eagle_nests %>% 
  pivot_longer(
    cols = c(`2007`, `2009`), 
    names_to = "year", 
    names_ptypes = list(year = integer()),
    values_to = "num_nests",
    values_ptypes = list(num_nests = integer())
  )

```


Q5/ Now, can you return the tidy Tibble to its original form using `pivot_wider()`?


```{r privot_wider, echo=FALSE}
example_eagle_nests_tidy %>% 
  pivot_wider(
    names_from = year, 
    values_from = num_nests
  )
```

## Gymnastics data

Now we will look at data of gymnatics scores from the 2012 and 2016 Olympics.

```{r gymdata}
example_gymnastics_2
```
Q6/ a/ Is this data tidy?

b/ What variables can you identify? (there should be four)

We can use `pivot_longer()` as before to create a dataframe with `country`, `event_year` and `score` columns.
```{r gymlong}
example_gymnastics_2 %>% 
  pivot_longer(
    cols = -country,
    names_to = "event_year",
    values_to = "score"
  )
```

However, `event_year` contains two variables, so the result is not yet tidy. Instead we want a seperate `event` variable and `year` variable.

Using `pivot_longer()`'s `names_sep` arguement, seperate these variables and make the `example_gymnatstics_2` tibble tidy!

```{r gymtidy, echo=FALSE}
example_gymnastics_2 %>% 
  pivot_longer(
    cols = -country,
    names_to = c("event", "year"),
    names_sep = "_",
    values_to = "score"
  )
```


## Extension exercise

Can you make the `example_gymnastics_3` and `example_gymnamtics_4` datasets tidy?


See [Data Challenge Lab](https://dcl-wrangle.stanford.edu/) for solutions.